Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2020-06-10 Moral-Communities-and-Crime/Moral_Communities_and_Crime.tex
v2 Working on revisions after Edivaldo N.

This concern is not idle: the endemic malady of violent crime represents
an important factor that threatens to derail sustainable development
goals. Violent crime, furthermore, tends to be more acute in those
regions most in need of development. On the one hand, violent crime
represents an economic and social drag (Becker and Kassouf 2017). The
impacts are multifaceted, since they affect people's well-being, either
through the loss of human life, mental health, and limitations in the
right to public spaces (Doran and Burgess 2011), or through disturbances
in schooling and academic achievement (United Nations Office on Drugs
and Crime 2019b). In turn, these negative effects combine to create the
unfortunate conditions that tend to breed criminal behavior, thus
creating a vicious cycle of economic disadvantage and violence (United
Nations Office on Drugs and Crime 2019b). It is not surprising, given
these peculiarities, that researchers have joined calls for increased
attention to the study of the patterns of violent crime in Low and
Middle Income Countries (LMIC; Murray, Cerqueira, and Kahn 2013).

Added to the scenario, formal social control institutions in countries
like Brazil leave something (or much) to be desired: while in practice
they should inhibit criminal behavior, they suffer from deep
deficiencies that end up reducing the deterrent power of the justice
system. Serious institutional problems include the inefficiency of the
police, the lack of national legislation, the glacial pace of judicial
processes, and the weak situation of the prison system in the country
(Menezes et al. 2013). This makes it even more urgent to understand the
role of \emph{informal social controls}, i.e., the ability of community
organizations such as schools, clubs, and neighborhood associations to
suppress crime by strengthening the capabilities of neighbors to control
inappropriate behavior (Groff 2015).

@@

With the above considerations in mind, the objective of this paper is to
investigate whether and how places of worship correlate spatially with
criminal events, or in other words, to investigate whether their signals
as moral communities have a discernible spatial dimension. The case
study is the city of Recife in the state of Pernambuco, in Brazil's
Northeast. Recife is a large and important metropolitan area in a
historically poor region, and afflicted by high levels of violent crime,
having the dubious honor of being one of five state capitals with the
highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

@@

\hypertarget{background}{%
\section{Background: Theoretical Perspectives on Religion and
Crime}\label{background}}

@@

Despite its intuitive appeal, research has not provided much direct
evidence for the hypothesis of hellfire (Hoffmann 2015, 1). In a bizarre
twist, even mainstream doctrines (such as absolution) may in fact
encourage beliefs that neutralize the fear of \emph{terrenal}
punishment, and thus turn out to be criminogenic (e.g., Topalli,
Brezina, and Bernhardt 2013). More deviant cases can even recast
criminal behavior as a form of spiritual insurgency (Chesnut 2017).
Counterexamples like this notwithstanding, researchers remain open to
the possibility that ``religion may still serve as a social control
mechanism by encouraging conventional beliefs, monitoring behaviors,
enhancing family attachments, or providing conventional activities''
(Hoffmann 2015, 1). In this way, the hypothesis of \emph{moral
communities} (Stark 1996) recognizes that social integration is
essential for increasing social control, thus reducing the practice of
behavior that is not in accordance with current norms. Rohrbaugh and
Jessor (1975) also emphasize that religion acts as social control since
it defines what is an appropriate attitude according to moral values,
thus making moral communities a close relative of the concept of
\emph{informal social control}, defined as ``the ability of social
groups or institutions to make norms or rules effective'' (Reiss 1951,
196; cited in Groff 2015, 91).

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plausibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization theory are
thought to occur at relatively small geographical scales.

@@

@@

The study is of the city of Recife, the capital of the state of
Pernambuco in the Northeast region of Brazil. With a population of
1,550,390 million inhabitants in 2010, Recife is one of the main
Brazilian metropolises, exerting a great economic influence in
neighboring regions. However, the city experiences a serious problem
with violent crime, and has the dubious honor of being one of the five
capital cities in Brazil with the highest homicide rates in the period
under study (Menezes et al. 2013).

@@

The data are organized at the individual level and it is possible to
obtain information about the location, day of the week, day of the
month, period of the day, as well as gender, age, and race of the
victim. The database used in this study comprises the period from July
1, 2008 to June 30, 2010. Some descriptive statistics regarding this
dataset are reported in Table 1, where it can be seen that overall,
about 91\% of the victims of LIVC in the period analyzed were men, while
approximately 98\% of violent crimes were homicides. In addition, most
of the victims were black or brown, and the youth population between the
ages of 18 and 30 is the most affected by violent crime. Lastly, it
should be noted that about 88\% of the criminal events under analysis
involved firearms. Figure \ref{fig:plot-crime} shows the spatial
distribution of the 1,657 LIVC crimes that occurred in the city of
Recife between July, 2008 and June, 2010.

@@

\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/fig-plot-crime-1} \hfill{}

@@

In addition to places of worship, the National Register of Addresses for
Statistical Purposes was queried to extract facilities other than places
of worship. As discussed above, the idea is to identify points of
reference that can be used as controls, having a neutral morality
profile. For the sake of the present study, we selected pharmacies, ice
cream shops, and bakeries to construct our control group. These three
types of establishments comply with the criteria of being morally
neutral and having a spatial distribution commensurate with places of
worship.

Figure \ref{fig:plot-events} shows the spatial distribution of the
places of worhsip and control establishments in the city of Recife. Note
the similarity between the maps. As expected, there are differences in
the number of points, but the locations of cases and controls are quite
similar.

@@

\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/fig-plot-events-1} \hfill{}

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 7954.19 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 0.59 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 0.63 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 0.70 & 0.17\\
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 0.21 & 0.38\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 53.91\% of the variance}\\
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-sed-as-quintiles-1.pdf}
\caption{\label{fig:plot-sed-as-quintiles}Socio-Economic Deprivation in
Recife classified by quintiles}
\end{figure}

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. In essence, this entails counting the number of
events at each level of the covariate, and then calculating an estimator
of the intensity.

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional
quadrat counts for Lethal and Intentional Violent Crime. The dotted line
indicates the global intensity, simply the total number of criminal
events divided by the area of the region. It can be seen that the
intensity of crime clearly increases with Socio-Economic Deprivation,
and the intensity in areas with the highest levels of Socio-Economic
Deprivation is 208.71\% higher than in places with the lowest levels of
Socio-Economic Deprivation. This is as expected, as research has
consistently found correlations between Socio-Economic Deprivation and a
variety of negative health outcomes, including homicide (Ichihara et al.
2018). A \(\chi^2\) test of independence on the quadrat counts yields a
value of \(p\leq 0.0001\) (four degrees of freedom), which comfortably
rejects the null hypothesis of homogeneity of intensity of crime by
level of Socio-Economic Deprivation.

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-crime-quadrat-1.pdf}
\caption{\label{fig:plot-crime-quadrat}Intensity of crime by level of
Socio-Economic Deprivation; the dotted line indicates the global
intensity of crime}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-events-quadrat-1.pdf}
\caption{\label{fig:plot-events-quadrat}Intensity of places of worship
and commercial establishments by level of Socio-Economic Deprivation;
the dotted lines indicates the respective global intensities}
\end{figure}

@@

\caption{\label{tab:table-quadrat-test-statistics}\label{tab:quadrat-test-statistics}$\chi^2$ tests of independence of conditional quadrat counts}
\centering
\begin{tabular}[t]{lccc}
\toprule
Event & Statistic & Degrees of Freedom & p-value\\
\midrule
Catholic & 6.913 & 4 & 0.2811\\
Evangelical & 38.669 & 4 & $<0.001$\\
Spiritist & 9.108 & 4 & 0.1169\\
Ice Cream Shop & 15.108 & 4 & 0.009\\
Pharmacy & 51.869 & 4 & $<0.001$\\
\addlinespace
Supermarket & 23.826 & 4 & $<0.001$\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/figure-plot-relative-distribution-1.pdf}
\caption{\label{fig:plot-relative-distribution}Intensity of crime as a
function of distance to a selected place of worship or commercial
establishment; bands are 95\% confidence interval and the dotted line
indicates the global intensity of crime}
\end{figure}

@@

The peak intensity as a function of distance to Evangelical churches is
3.33 times the baseline intensity; for Catholic places of worship, the
highest intensity is 2.12 times the baseline intensity; and the
intensity reaches a peak of 2.67 times the baseline intensity when the
covariate is distance to Spiritist churches. Turning now to inspect the
three types of commercial establishments, we see that the highest
intensities are 2.28, 2.46, and 2.09 times the intensity according to
Socio-Economic Deprivation for ice cream shops, pharmacies, and
supermarkets, respectively.

Some differences are observed as well. Whereas the peak intensity as a
function of distance to Evangelical places of worship was 183.78\%
higher than the intensity of Catholic places of worship \emph{without a
baseline function}, once that we account for Socio-Economic Deprivation
by means of a baseline function, the peak relative intensity as a
function of distance to Evangelical churches declines to 176.53\% higher
than the intensity as a function of distance to Catholic places of
worship at the same distance. Accounting for Socio-Economic Deprivation
in this analysis of intensity has in general a moderating effect:
whereas the peak intensity of crime was was 194.54\% higher when
comparing Evangelical and Spiritist churches at the distance of the peak
intensity of Evangelical churches, this difference declines to 186.45\%
after accounting for Socio-Economic Deprivation. Similarly, this changes
from 174.19\% to 168.33\% in the case of ice cream shops, from 151.52\%
to 140.97\% in the case of pharmacies, and finally from 177.63\% to
164.33\% in the case of supermarkets.

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/figure-plot-relative-distribution-with-baseline-1.pdf}
\caption{\label{fig:plot-relative-distribution-with-baseline}Intensity
of crime as a function of distance to a selected place of worship or
commercial establishment after introducing Socio-Economic Deprivation as
a baseline function; bands are 95\% confidence intervals and the dotted
line indicates the baseline intensity with respect to Socio-Economic
Deprivation}
\end{figure}

@@

>>>>

This concern is not idle: violent crime tends to be more acute in those
regions most in need of development. On the one hand, violent crime
represents an economic and social drag (Becker and Kassouf 2017). The
impacts are multifaceted, since they affect people's well-being, either
through the loss of human life, mental health, and limitations in the
right to public spaces (Doran and Burgess 2011), or through disturbances
in schooling and academic achievement (United Nations Office on Drugs
and Crime 2019b). In turn, these negative effects combine to create the
unfortunate conditions that tend to breed criminal behavior, thus
creating a vicious cycle of economic disadvantage and violence (United
Nations Office on Drugs and Crime 2019b). It is not surprising, given
these peculiarities, that researchers have joined calls for increased
attention to the study of the patterns of violent crime in Low and
Middle Income Countries (LMIC; Murray, Cerqueira, and Kahn 2013).

Added to the scenario, formal social control institutions in countries
like Brazil often leave something (or much) to be desired: while in
practice they should inhibit criminal behavior, they suffer from deep
deficiencies that end up eroding the deterrent power of the justice
system. Serious institutional problems include the inefficiency of the
police, the lack of national legislation, the glacial pace of judicial
processes, and the weak situation of the prison system in the country
(Menezes et al. 2013). This makes it even more urgent to understand the
role of \emph{informal social controls}, i.e., the ability of community
organizations such as schools, clubs, and neighborhood associations to
suppress crime by strengthening the capabilities of neighbors to control
inappropriate behavior (Groff 2015).

@@

With the above considerations in mind, the objective of this paper is to
investigate whether and how places of worship correlate spatially with
criminal events, or in other words, to investigate whether their signals
as moral communities have a discernible spatial dimension. Our argument
is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Since places of worship are associated with normative codes of
  behavior, can they by their presence contribute to create moral
  communities? The mechanisms for this are explained by Groff (Groff
  2015) in terms of ``Eyes on the Street'' (see Jacobs 1961, 1968;
  Appleyard 1980) and Human Territorial Functioning (Taylor 1988, 1998).
\item
  If so, what is the distance decay of said moral community effect? Is
  it uniform over a certain distance or does it vary?
\end{enumerate}

The case study is the city of Recife in the state of Pernambuco, in
Brazil's Northeast. Recife is a large and important metropolitan area in
a historically poor region, and afflicted by high levels of violent
crime, having the dubious honor of being one of five state capitals with
the highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

@@

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{theoretical-perspectives-on-religion-and-crime}{%
\subsection{Theoretical Perspectives on Religion and
Crime}\label{theoretical-perspectives-on-religion-and-crime}}

@@

Despite its intuitive appeal, research has not provided much direct
evidence for the hypothesis of hellfire (Hoffmann 2015, 1). In a bizarre
twist, even mainstream doctrines (such as absolution) may in fact
encourage beliefs that neutralize the fear of \emph{terrenal}
punishment, and thus turn out to be criminogenic (e.g., Topalli,
Brezina, and Bernhardt 2013). More deviant cases can even recast
criminal behavior as a form of spiritual insurgency (Chesnut 2017).
Counterexamples like this notwithstanding, researchers remain open to
the possibility that ``religion may still serve as a social control
mechanism by encouraging conventional beliefs, monitoring behaviors,
enhancing family attachments, or providing conventional activities''
(Hoffmann 2015, 1). In this way, the hypothesis of \emph{moral
communities} (Stark 1996) recognizes that social integration is
essential for increasing social control, thus reducing the practice of
behavior that is not in accordance with current norms. Rohrbaugh and
Jessor (1975) also emphasize that religion acts as social control since
it defines what is an appropriate attitude according to moral values,
thus making moral communities a close relative of the concept of
\emph{informal social control}, defined as ``the ability of social
groups or institutions to make norms or rules
effective''\footnote{The discussion about disorder is directly related to the idea of social control. According to this perspective, control mechanisms to contain deviant actions are considered indispensable [@Rodrigues2012medo]. This theory assumes that the deviant act is seen as normal in social organization, and coercive forces are needed to limit such impulses. This control can be exercised through formal institutions, as well as through informal mechanisms. It would be a set of norms as well as positive and negative sanctions, which are specified throughout the socialization process, which aims to establish values and build the individual's morals, disciplining society and subjecting individuals to standards and principles predefined. As the individual finds himself more and more inserted in the social context, less will be the chances of criminal behavior. From this perspective, the existence of bonds of solidarity and trust would be relevant for the effective exercise of supervision over crime in the neighborhood. Informal control is then linked to the ability of a neighborhood to establish communication and surveillance networks, interfering in cases of criminal actions and increasing security in the region.}
(Reiss 1951, 196; cited in Groff 2015, 91).

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plausibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization
theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.}
are thought to occur at relatively small geographical scales.

@@

\hypertarget{religion-in-brazil}{%
\subsection{Religion in Brazil}\label{religion-in-brazil}}

Brazil is a country with a Christian predominance. In the first national
census in which religion was considered, in 1872, practically the entire
population (99.72\%) identified themselves as Catholic. In the 1940
census, when the religion question was researched again, Catholicism was
still the religion with which the vast majority of the population
identified: 95.2\% of Brazilians declared themselves as Catholics that
year.

The dominance of Catholicism in the religious life of Brazil is
unsurprising given the colonial past of the country. However, the
religious makeup of the population started was not completely
homogeneous. Among other offerings in the market place of spiritual
ideas, Spiritism stands out. The Spiritist Doctrine was founded in
nineteenth-century France by Allan Kardec as a way of understanding the
world and its relations with the ``beyond'': a curious mix of
pseudo-scientific language, philosophy, and religion (Arribas 2011a).
Spiritism arrived in Brazil in the middle of the previous century and
since then it slowly spread in large and medium-sized cities. Today,
Spiritism is the third largest religious segment in Brazil, after
Catholicism and Evangelical Christianism, and shares with these
religions the Christian principle of charity through pious assistance.

Another signal event in the developmento of religion in Brazil is the
arrival of Pentecostal Protestant churches, such as the Christian
Congregation of Brazil in 1910 and the Assembly of God in 1911.
Initially, these religious denominations did not make important inroads
in Brazilian society (as evidenced by the 1940 census). However, in the
context of rapid industrialization and major urbanization, they
contributed to the change in the religious landscape along with other
religions (Souza 2019): according to the 2010 census the configuration
of the religious scene in Brazil had changed considerably from the
previous century: 64.6\% of the population identified as Catholic and
22.2\% as Evangelical Christians. The acceleration of this demographic
change was directly related to the inroads made by Evangelical churches:
The Assembly of God, for instance, grew considerably to total more than
12.3 million adherents. This church is represented extensively in many
regions of the country, following a system of loosely allied ministries.
With a charismatic profile and the lack of a centralized authority,
growth of this church has occurred mainly through subdivision into
numerous small and autonomous churches. These churches have in common
the Prosperity Gospel and an emphasis on economic entrepreneurship
(e.g., Mora 2008). They also do not shy away from partisan political
action.

In general, there have been two major changes in Brazil regarding
religious issues: the slow decline of Catholicism and the rapid growth
of Evangelical Christianism. These changes are evident in the daily life
of the average Brazilian, either through the radio or through TV
programs, which give more and more space to religious doctrines. At the
same time, there has been an increase in cases of religious intolerance
in Brazil, generating violence acts against temples, objects of worship,
and people.

@@

The study is of the city of Recife, the capital of the state of
Pernambuco in the Northeast region of Brazil. With a population of
1,550,390 million inhabitants in 2010, Recife is one of the main
Brazilian metropolises, exerting a great economic influence in
neighboring regions. However, the city experiences a serious problem
with violent crime, and has the dubious honor of being one of the five
capital cities in Brazil with the highest homicide rates in the period
under
study\footnote{The appendix included information that allows a more in-depth characterization of the city of Recife: population distribution, income, inequality, urban facilities (education and health) and location of residential areas.}
(Menezes et al. 2013).

@@

The data are organized at the individual level and it is possible to
obtain information about the crime event location, day of the week, day
of the month, period of the day, as well as gender, age, and race of the
victim. The database used in this study comprises the period from July
1, 2008 to June 30, 2010. Some descriptive statistics regarding this
dataset are reported in Table 1, where it can be seen that overall,
about 91\% of the victims of LIVC in the period analyzed were men, while
approximately 98\% of violent crimes were homicides. In addition, most
of the victims were black or brown, and the youth population between the
ages of 18 and 30 is the most affected by violent crime. Lastly, it
should be noted that about 88\% of the criminal events under analysis
involved firearms. Figure \ref{fig:plot-crime} shows the spatial
distribution of the 1,657 LIVC crimes that occurred in the city of
Recife between July, 2008 and June, 2010.

@@

\includegraphics{Moral_Communities_and_Crime_files/figure-latex/fig-plot-crime-1} \hfill{}

@@

In addition to places of worship, the National Register of Addresses for
Statistical Purposes was queried to extract facilities other than places
of worship. To study the spatial relationship between churches and
crime, a control group was built and then a comparison of the empirical
distribution was carried out. Gilberston and Hayes (2012) highlight that
different configurations of urban space affect how people perceive the
environment around them. In this perspective, churches are continually
affecting the impression that people have of a region. Thus, we assume
that churches have unique characteristics that may or may not inhibit
the occurrence of a criminal act: they are urban constructions that are
not neutral in relation to crime. On the other hand, there are urban
facilities that do not provide any specific reason to justify the fact
that the crime occurs closer or further from these buildings. Thus, it
can be thought that any urban construction belongs to one of the
following distinct situations: neutral or non-neutral to crime.
Pharmacies, ice cream parlors and bakeries will be used in the
construction of the control group and will be considered neutral. The
comparison between the distributions of non-neutral and crime-neutral
establishments forms the empirical basis of the study. The choice of
establishments that make up the control was based on two criteria: (a)
being neutral to crime and (b) having a spatial distribution similar to
non-neutral establishments. The strategy used is to compare similar
establishments about their spatial distribution, but different in
relation to their neutrality profile in relation to crime, seeking to
simulate the existence of a case and a control group. The case will be
composed of churches, while the control will be pharmacies, ice cream
parlors and bakeries.

As discussed above, the idea is to identify points of reference that can
be used as controls, having a neutral morality profile. For the sake of
the present study, we selected pharmacies, ice cream shops, and bakeries
to construct our control group. These three types of establishments
comply with the criteria of being morally neutral and having a spatial
distribution commensurate with places of worship.

Figure \ref{fig:plot-events} shows the spatial distribution of the
places of worhsip and control establishments in the city of Recife. Note
the similarity between the maps. As expected, there are differences in
the number of points, but the locations of cases and controls are quite
similar: the distribution of the churches, as well as of the commercial
establishments follow the same pattern, being correlated with the
spatial distribution of the population (see in the appendix). Since such
urban constructions respond to society's demand, they tend to be located
close to population agglomerations.

@@

\includegraphics{Moral_Communities_and_Crime_files/figure-latex/fig-plot-events-1} \hfill{}

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Population Density & 0.00 & 10564.16 & 16333.62 & 19230.23 & 25106.07 & 181642.55 & 0.03\\
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 7954.19 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 0.59 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 0.63 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 0.70 & 0.17\\
\addlinespace
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 0.21 & 0.38\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 53.91\% of the variance}\\
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_files/figure-latex/plot-sed-as-quintiles-1.pdf}
\caption{\label{fig:plot-sed-as-quintiles}Socio-Economic Deprivation in
Recife classified by quintiles}
\end{figure}

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. In essence, this entails counting the number of
events at each level of the covariate, and then calculating an estimator
of the intensity. The global intensity of each type of event (i.e.,
places of worship and businesses), as well as quadrant counts and
intensity at each level of SED can be consulted in Table
\ref{tab:quadrat-count-results}.

\caption{\label{tab:table-counts-intensity-summary}\label{tab:quadrat-count-results}Conditional quadrat counts results}
\centering
\begin{tabular}[t]{lcccl}
\toprule
Event & Global Intensity & SED Quintile & Quadrat Count & Intensity\\
\midrule
 &  & Low & 201 & 0\\

 &  & Middle Low & 295 & 0\\

 &  & Middle & 364 & 0\\

 &  & Middle High & 479 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Violent Crime} &  & High & 318 & 0\\

 &  & Low & 32 & 0\\

 &  & Middle Low & 26 & 0\\

 &  & Middle & 33 & 0\\

 &  & Middle High & 29 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Catholic} &  & High & 11 & 0\\

 &  & Low & 209 & 0\\

 &  & Middle Low & 304 & 0\\

 &  & Middle & 334 & 0\\

 &  & Middle High & 345 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Evangelical} &  & High & 185 & 0\\

 &  & Low & 37 & 0\\

 &  & Middle Low & 50 & 0\\

 &  & Middle & 40 & 0\\

 &  & Middle High & 40 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Spiritist} &  & High & 20 & 0\\

 &  & Low & 45 & 0\\

 &  & Middle Low & 60 & 0\\

 &  & Middle & 39 & 0\\

 &  & Middle High & 37 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Ice Cream} &  & High & 26 & 0\\

 &  & Low & 116 & 0\\

 &  & Middle Low & 122 & 0\\

 &  & Middle & 91 & 0\\

 &  & Middle High & 84 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Pharmacies} &  & High & 25 & 0\\

 &  & Low & 27 & 0\\

 &  & Middle Low & 32 & 0\\

 &  & Middle & 38 & 0\\

 &  & Middle High & 16 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Super Markets} & \multirow{-35}{*}{\centering\arraybackslash 0} & High & 12 & 0\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional
quadrat counts for Lethal and Intentional Violent Crime. The dotted line
indicates the global intensity, simply the total number of criminal
events divided by the area of the region. It can be seen that the
intensity of crime clearly increases with Socio-Economic Deprivation,
and the intensity in areas with the highest levels of Socio-Economic
Deprivation is 216.2\% higher than in places with the lowest levels of
Socio-Economic Deprivation. This is as expected, as research has
consistently found correlations between Socio-Economic Deprivation and a
variety of negative health outcomes, including homicide (Ichihara et al.
2018). A \(\chi^2\) test of independence on the quadrat counts yields a
value of \(p\leq 0.0001\) (four degrees of freedom), which comfortably
rejects the null hypothesis of homogeneity of intensity of crime by
level of Socio-Economic Deprivation.

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_files/figure-latex/plot-crime-quadrat-1.pdf}
\caption{\label{fig:plot-crime-quadrat}Intensity of crime by level of
Socio-Economic Deprivation; the dotted line indicates the global
intensity of crime}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_files/figure-latex/plot-events-quadrat-1.pdf}
\caption{\label{fig:plot-events-quadrat}Intensity of places of worship
and commercial establishments by level of Socio-Economic Deprivation;
the dotted lines indicates the respective global intensities}
\end{figure}

@@

\caption{\label{tab:table-quadrat-test-statistics}\label{tab:quadrat-test-statistics}$\chi^2$ tests of independence of conditional quadrat counts}
\centering
\begin{tabular}[t]{lccc}
\toprule
Event & Statistic & Degrees of Freedom & p-value\\
\midrule
Catholic & 7.074 & 4 & 0.264\\
Evangelical & 32.976 & 4 & $<0.001$\\
Spiritist & 7.561 & 4 & 0.2181\\
Ice Cream Shop & 13.946 & 4 & 0.0149\\
Pharmacy & 53.220 & 4 & $<0.001$\\
\addlinespace
Supermarket & 18.819 & 4 & 0.0017\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_files/figure-latex/figure-plot-relative-distribution-1.pdf}
\caption{\label{fig:plot-relative-distribution}Intensity of crime as a
function of distance to a selected place of worship or commercial
establishment; bands are 95\% confidence interval and the dotted line
indicates the global intensity of crime}
\end{figure}

@@

The peak intensity as a function of distance to Evangelical churches is
3.32 times the baseline intensity; for Catholic places of worship, the
highest intensity is 2.12 times the baseline intensity; and the
intensity reaches a peak of 2.67 times the baseline intensity when the
covariate is distance to Spiritist churches. Turning now to inspect the
three types of commercial establishments, we see that the highest
intensities are 2.28, 2.45, and 2.09 times the intensity according to
Socio-Economic Deprivation for ice cream shops, pharmacies, and
supermarkets, respectively.

Some differences are observed as well. Whereas the peak intensity as a
function of distance to Evangelical places of worship was 183.78\%
higher than the intensity of Catholic places of worship \emph{without a
baseline function}, once that we account for Socio-Economic Deprivation
by means of a baseline function, the peak relative intensity as a
function of distance to Evangelical churches declines to 176.48\% higher
than the intensity as a function of distance to Catholic places of
worship at the same distance. Accounting for Socio-Economic Deprivation
in this analysis of intensity has in general a moderating effect:
whereas the peak intensity of crime was was 194.54\% higher when
comparing Evangelical and Spiritist churches at the distance of the peak
intensity of Evangelical churches, this difference declines to 186.43\%
after accounting for Socio-Economic Deprivation. Similarly, this changes
from 174.19\% to 168.37\% in the case of ice cream shops, from 151.52\%
to 140.98\% in the case of pharmacies, and finally from 177.63\% to
164.31\% in the case of supermarkets.

@@

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_files/figure-latex/figure-plot-relative-distribution-with-baseline-1.pdf}
\caption{\label{fig:plot-relative-distribution-with-baseline}Intensity
of crime as a function of distance to a selected place of worship or
commercial establishment after introducing Socio-Economic Deprivation as
a baseline function; bands are 95\% confidence intervals and the dotted
line indicates the baseline intensity with respect to Socio-Economic
Deprivation}
\end{figure}

@@

\leavevmode\hypertarget{ref-Appleyard1980livable}{}%
Appleyard, D. 1980. ``Livable Streets - Protected Neighborhoods.''
Journal Article. \emph{Annals of the American Academy of Political and
Social Science} 451 (SEP): 106--17.
\url{\%3CGo\%20to\%20ISI\%3E://A1980KK14700011}.
